---
title: "Custom Data-Driven Colors in ggplot2"
format: hugo-md
date: 2026-06-09
editor: visual
---

Obviously you'll need ggplot2 installed in your R environment to do this. I've also generated some bogus data just to keep things simple for this.

```{r}
library(ggplot2)
FullData <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David"),
  score = c(42, 37, 91, 58),
  food = c("pizza", "tacos", "pizza", "burger")
)
print(FullData)
```

The easy way to generate a basic bar chart is as follows, which most of you should know by now.

```{r}
ggplot(data=FullData, aes(x=name, y=score))+
  geom_col()
```

Now let's say I want to color-code them to each type of food, with the following order: blue for pizza, red for tacos, and green for burger.

The easiest way to accomplish that is simply by manually assigning it in the "fill" field by combining them all into a 1D list from left to right, as I have done below.

```{r}
ggplot(data=FullData, aes(x=name, y=score))+
  geom_col(fill=c("blue", "red", "blue", "green"))
```

Now this does work, but for one there just isn't a key on this graph, so viewers have no idea what the color means in the first place. Also, since the color hard-coded, if any one person's food changes or we insert someone else in the middle, we have to manually adjust the list of colors.

Instead, we can use the aesthetic mapping, which tells ggplot how to use data-driven components to describe your graph, specifically for the "fill" value.

```{r}
ggplot(data=FullData, aes(x=name, y=score, fill=food))+
  geom_col()
```

Now this fixes the "no key" and "need to manually update colors" problems, but now the colors are randomly chosen by ggplot instead of by us!

To set them ourselves, we can use the "scale_fill_manual" tool to give it a special mapping of what colors we want to be assigned to each type of food.

```{r}
ColorCode = c("pizza" = "blue", "tacos" = "red", "burger" = "green")
ggplot(data=FullData, aes(x=name, y=score, fill=food))+
  geom_col()+
  scale_fill_manual(values=ColorCode)
```

This is about as good as we can get it to be. Of course if this were to be presented, you should add a chart title, make the axes have properly formatted names, and do some light theming, but the core of these features can be applied to any graph you want to!
